Your time series is (probably) too short

### 2017-08-14 writing/manuscript_draft.Rmd
added stacked barplot for density plot

---
bibliography: White_bib.bib
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{setspace}\doublespacing \usepackage{lineno}
  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers

@@

To be submitted to: <!--\emph{Biological Conservation} (Short Communication)-->
\vspace{3 mm}

@@

# Abstract
Long-term census data are necessary to better understand ecological processes and make management decisions. However, census data are often expensive, requiring a lot of people-hours and equipment. Little work has addressed the question, when is a time series of census data long enough to address a question of interest? Here, we explore two approaches to address this question: simulations and an empirical approach. We specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability. In addition, we examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. <!--Here, We stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.

@@

Naturally, one might ask, how long of a time series is actually necessary? This question has important implications for both research and management. Scientifically, we need to know the time series length required to address a specific question. Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations). From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not. The IUCN Red List Categories and Criteria suggest under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we do not want to sample for longer than is necessary [@Gerber1999].

@@

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on single species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions. More specifically, we estimate the minimum time series length required to assess long-term changes in abundance via simple linear regression. We examine 878 population time series to estimate this minimum time series length required for detecting population trends.

@@

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Johnson2015] (and cite Bolker). We only briefly discuss this approach as it has been described elsewhere (citations). Essentially, we take a stochastic population model and simulate it for a number of years repetitively. We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question. This approach requires us to determine values for our model parameters (e.g. growth rate). As an example, we can take the following population model for population size $N$ at time $t$:
$$
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
$$
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of increase or decrease.

By simulating this model repetitively for different lengths of time, we can determine the minimum time series required. Here statistical power is simply the fraction of simulations with significant trends in abundance. A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression. Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8, @Cohen1992). Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.

In Fig. \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods ($t=40$). It is clear that statistical power increases quickly with increases in length of time sampled (Fig. \ref{fig:theoretical_approach}b). Once power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$).

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data are originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015] like body size and generation length. All `r nrow(LPI_pop_info)` populations are within the Aves class. In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999; @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true'', time series also has a significant slope. This is a measure of statistical power (often denoted as $1 - \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention [@Cohen1992].

@@

We determined the minimum time series length required to address a particular question of interest. More specifically we asked, what is the minimum time series length required to determine, via linear regression, the long-term trend in population abundance? The minimum time series length required is that which had high enough statistical power. We investigated this question through two approaches: simulation-based and empirical analyses. We briefly describe results from the simulation approach and then discuss our empirical approach.

@@

We constructed a general population model where the trend over time could be a model parameter. We chose a simple model, but any other population model could be used. The specific model choice should be tailored to the population of interest. We then ran simulated time series of different lengths. From these simulations we determined the minimum time series length required to achieve a certain level of power (which we set to 0.8). We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$.

In line with past work, we found the $T_{min}$ increases (i.e. more time is required) with decreases in trend strength and with increases in variability (citation).

## Empirical test

@@

plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')
mtext(text = 'Frequency',2,line=3,cex=1.2)
mtext(text = 'Minimum time required',1,line=3,cex=1.2)
```

@@

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}). Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Fig. SX). Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.

We also examined a subset of populations where more explanatory variables were known (e.g. body mass, longevity, location). None of these explanatory variables were predictive of the minimum time series length required (Fig. \ref{fig:biological_correlates}). Some differences if $T_{min}$ did occur between biological class (Fig. \ref{fig:class}), but this can be attributed to the amount populations are known to fluctuate (i.e. the variability in population size).

@@

We explored two different approaches to estimate the minimum time series length required to address a particular question of interest. We specifically asked, what is the minimum time series length required to determine long-term changes in abundance using linear regression? The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].

@@

# Figures

stuff here \ref{fig:empirical_approach_example}

>>>>

---
bibliography: White_bib.bib
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{setspace}\doublespacing \usepackage{lineno} \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}

@@

To be submitted to: \emph{Biological Conservation} (Short Communication)
\vspace{3 mm}

@@

# Abstract
Long-term census data are necessary to better understand ecological processes and make management decisions. However, census data are often expensive, requiring a lot of time and resource. Little work has addressed the question, when is a time series of census data long enough to address a question of interest? Here, we explore two approaches to address this question: simulations and an empirical approach. We specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability. In addition, we examine 878 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends. Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.

@@

Naturally, one might ask, how long of a time series is actually necessary? This question has important implications for both research and management. Scientifically, we need to know the time series length required to address a specific question. Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations). From a management perspective, it is important to understand when population trend over time is actually meaningful or not. The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary [@Gerber1999].

@@

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address several questions. In the main manuscript, we estimate the minimum time series length required to assess long-term changes in abundance via simple linear regression. We then examine 878 population time series to estimate this minimum time series length required for detecting population trends. We expand on the results in the supplmentary material by examining other questions of interest.

@@

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015]. We only briefly discuss this approach as it has been described elsewhere. Essentially, we use a population model and simulate it for a number of years repetitively. We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question. This approach requires us to determine values for our model parameters (e.g. population variability). As an example, we can take the following population model for population size $N$ at time $t$:
$$
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
$$
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of the increase or decrease (i.e. the rate of increase).

By simulating this model repetitively for different lengths of time, we can determine the minimum time series required. Here, statistical power is simply the fraction of simulations with significant trends in abundance. A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression. Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8, @Cohen1992). Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.

In Fig. \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods ($t=40$). It is clear that statistical power increases quickly with increases in length of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$).

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data are originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. <!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015], including body size and generation length. All `r nrow(LPI_pop_info)` populations are within the Aves class. In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999; @Giron-Nava2017]. Then, we determine the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true'', time series also has a significant slope. This is a measure of statistical power (often denoted as $1 - \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention [@Cohen1992].

@@

We determined the minimum time series length ($T_{min}$) required to address a particular question of interest. More specifically we asked, what is the minimum time series length required to determine, via linear regression, the long-term population trend? Here, the minimum time series length required had high enough statistical power (greater than 0.8) for a set significance level of 0.05. We investigated this question through two approaches: simulation-based and empirical analyses. We briefly describe results from the simulation approach and then discuss our empirical approach.

@@

We constructed a general population model where the trend (i.e. slope coefficient) over time could be a model parameter. We chose a simple model, but any other population model could be used (Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice should be tailored to the population of interest. We then ran simulated time series of different lengths. From these simulations we determined the minimum time series length required to achieve a certain level of statistical power. We were also able to alter the parameter values (trend strength and population variability) of the population model to see how this altered $T_{min}$ (Fig. \ref{fig:theoretical_approach}).

In line with past work [@Gerrodette1987], we found the $T_{min}$ increases (i.e. more time is required) with decreases in trend strength and with increases in variability (Fig. \ref{fig:theoretical_approach}).

@@

#plot(density(pop_info$min_time_for_power),las=1,ylab='',xlab='',cex.axis=1.2,main='')

counts <- table(pop_info$class,pop_info$min_time_for_power)
barplot(counts, main=" ",
        xlab=" ", ylab="", col=1:4,
        legend = rownames(counts),ylim=c(0,50),xlim=c(0,38),
        las=1,cex.names =1.2,cex.axis = 1.2,density=30,
        angle=c(20,40,60,80))

mtext(text = 'Frequency',2,line=3,cex=1.2)
mtext(text = 'Minimum time required',1,line=3,cex=1.2)
```

@@

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as expected based on our simulation results and those of others [@Rhodes2011], the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}). Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table S1). Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.

We also examined a subset of populations where more explanatory variables were known (e.g. body mass, longevity, location). None of these explanatory variables were predictive of the minimum time series length required (Fig. \ref{fig:biological_correlates}). Some differences in $T_{min}$ did occur between biological class (Fig. \ref{fig:class}), but this can be attributed to the amount populations are known to fluctuate (i.e. the variability in population size).

@@

We explored two different approaches to estimate the minimum time series length required to address a particular question of interest. We specifically asked, what is the minimum time series length required to determine long-term population trends using linear regression? This is one of the simplest questions one could ask of a time series. We would expect more involved questions (e.g. detecting chaos) to require a larger minimum time series length. The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].

@@
